---
title: "GLM Model-Fit Forensics"
author: "STAT 637"
date: "September 21, 2026"
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

Dr. Berrett used Codex GPT-5.6 Luna to create this initial file and to suggest unique data sets for each team.  She then did extensive editing (sometimes using Claude Sonnet 5 in Positron for generalizing formulation) and checking to makes sure the activity examined the content in the way the course needs. 

## Activity goal

Act as a statistical consultant deciding whether a predictor should remain in a generalized linear model. Use the assigned data set to compare two nested models and deliver a one-minute model-fit verdict.

Your report should show how coefficient uncertainty, deviance comparisons, hypothesis tests, and residual diagnostics support the decision.

## Partnership assignments

| Pair | Partners | Assigned data set | Response | Candidate predictor |
|---:|---|---|---|---|
| 1 | Jarom and Andrew | `MASS::birthwt` | `low` | `smoke` |
| 2 | Anna and Brylee | `datasets::InsectSprays` | `count` | `spray` |
| 3 | Nathan and Shae | `datasets::esoph` | `cbind(ncases, ncontrols)` | `tobgp` |
| 4 | RJ and Brett | `MASS::quine` | `Days` | `Eth` |
| 5 | Tyler and Caleb | `datasets::mtcars` | `mpg` | `wt` |
| 6 | Bracken and Dallin | `MASS::Insurance` | `Claims` | `Group` |

All assigned data sets are available directly from R or the standard `MASS` package. No local teaching-folder files are required.

## Workflow

### Identify the modeling question

Discuss: description of the response variable, the scientific or practical question, and the predictor whose contribution you will investigate.

```{r datasetup}
# Replace these placeholders with your partnership's assigned values from the table.
data_obj <- MASS::quine
response <- "Days"
candidate_predictor <- "Eth"
fam <- "poisson"
```

### Prepare the data

Document recoding, missing-value handling, filtering, and any transformations. Explain why the selected response distribution is reasonable.

```{r}
# Example only. Adapt to your data set.
library(MASS)
dat <- MASS::quine

```

### Fit two nested models

Fit a reduced model without the candidate predictor and a full model that includes it. Keep the response distribution and link function the same when comparing the models.

```{r mods}
library(stats)

# reformulate() builds the formula from the *values* of response and
# candidate_predictor, rather than using those names literally.
mod0 <- glm(
  reformulate("1", response = response),
  data = dat,
  family = fam
)

mod1 <- glm(
  reformulate(candidate_predictor, response = response),
  data = dat,
  family = fam
)

summary(mod0)
summary(mod1)
```

### Quantify coefficient uncertainty

For the candidate coefficient, report the estimate, standard error, covariance information, and a confidence interval. Explain what the interval says in the scale of the linear predictor or response model.

```{r}
coef(summary(mod1))
vcov(mod1)
summary(mod1)$cov.scaled
confint(mod1)
```

Record:

- candidate coefficient estimate:
- standard error:
- relevant covariance entry or matrix:
- confidence intervals:
- coefficient interpretation:

### Compare model fit with deviance

Compute the difference in deviance and use it to test whether the full model improves fit relative to the reduced model.

```{r}
deviance(mod0)
deviance(mod1)
delta_deviance <- deviance(mod0) - deviance(mod1)
delta_deviance

anova(mod0, mod1)
```

State the null and alternative hypotheses, test statistic, degrees of freedom, p-value, and conclusion in context.

### Compare residuals

Compare at least two residual types. Include standardized residuals and identify observations that deserve investigation.

```{r}
residual_summary <- data.frame(
  fitted = fitted(mod1),
  raw = residuals(mod1, type = "response"),
  pearson = residuals(mod1, type = "pearson"),
  deviance = residuals(mod1, type = "deviance"),
  standardized = rstandard(mod1)
)

head(residual_summary)

par(mfrow=c(2,2))
plot(mod1)
```

Very briefly, discuss:

- why multiple residual types are available for GLMs;
- why standardization is useful;
- whether the residuals reveal lack of fit, unusual observations, or influential cases.

```{r}
dat$Days[c(59, 72, 104)]
```

## One-minute model-fit verdict

Use this structure for your team's one-minute presentation:

1. **Question:** What were you trying to learn?
Whether or not a student's ethnicity has a significant impact on modeling absenteeism
2. **Models:** What differed between the reduced and full models?
Reduced: only intercept
Full: Included Ethnicity
3. **Evidence:** What did the coefficient interval, deviance test, and residuals show?
Interval and Deviance test both showed that including ehtnicity has a significant impact on the model performance.
Residuals showed there may be some influential points, not a lot to learn due to binary input variabel. 
4. **Verdict:** Keep the candidate predictor, remove it, or gather better data. Explain why.
Keep because the deviance test showed that the models performed significantly differently, implying the 'full' model did significantly better and we should keep the variable in the model. 
Ethnicity A appears to 'absenteeism' more. 

Due to very high variance, running it as a negative binomial instead of a Poisson would likely be a better approach (Brett ran it and the residuals look much better - smaller range, etc. )

